Data Prep: IUCN and AquaMaps

### 2017-03-12 data_explore/data_prep.Rmd
updated the matching algorithm (again) and reran analysis

@@

@@

spp_all <- spp_iucn_maps %>%
  full_join(spp_am_maps, by = 'sciname') %>%
  distinct()

### some confusion due to non-matching scinames and IDs...
spp_all <- spp_all %>%
  mutate(iucn_match = iucn_sid == iucn_id,
         iucn_wrong_map = ifelse(!iucn_match & iucn_id %in% spp_all$iucn_sid, TRUE, FALSE)) %>%
  mutate(iucn_sid = ifelse(iucn_wrong_map & !is.na(iucn_wrong_map), iucn_id, iucn_sid)) %>%
  select(-iucn_match, -iucn_wrong_map, -iucn_id, -sciname_count) %>%
  distinct()

length(unique(x$iucn_sid)) ###  4013, incl NA
length(unique(x$am_sid))   ### 22890, incl NA

@@

# length(unique(spp_am_v_iucn$sciname)); length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))
### NOTE: 2314 unique scientific names; but 2314 unique am_sid and 2320 unique iucn_sid.
### Seasnakes are included in Reptiles (IUCN).  Two "mangroves" that are actually seagrasses.
spp_am_v_iucn <- spp_am_v_iucn %>%
  group_by(am_sid, iucn_sid) %>%
  mutate(dupes = n() - 1) %>%
  ungroup() %>%
  filter(!(dupes == 1 & spp_group == 'MANGROVES')) %>%
  filter(spp_group != 'SEASNAKES') %>%
  select(-dupes)
# length(unique(spp_am_v_iucn$sciname)); length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))

@@

spp_am_v_iucn <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))

@@

# iucn_spp_cells_files <- list.files(file.path(dir_anx, 'v2016/iucn_intersections'), full.names = TRUE)
# iucn_spp_cells_all <- lapply(iucn_spp_cells_files, FUN = function(x) {
#   read_csv(x) %>% 
#     mutate(id_no = as.integer(id_no)) %>%
#     rename(iucn_sid = id_no)
#   }) %>%
#   bind_rows()
# 
# iucn_spp_cells <- iucn_spp_cells_all %>%
#   filter(iucn_sid %in% spp_am_v_iucn$iucn_sid) %>%
#   filter(presence != 5)

### length(unique(iucn_spp_cells$iucn_sid)) results in 2271 species found...
###   some species with no cells reported.  
### spp_missing <- spp_am_v_iucn %>% filter(!iucn_sid %in% iucn_spp_cells$iucn_sid)
###   49 spp with no cells reported.  Shapefile errors?
# write_csv(iucn_spp_cells, file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'))

message('Loading AquaMaps species to cell table.  This might take a few minutes.')
am_spp_cells_all <- fread(file.path(dir_data_remote, 'aquamaps/d2015/csv/hcaf_sp_native_trunc.csv'), 
                         showProgress = TRUE)
am_spp_cells <- am_spp_cells_all %>%
  rename(am_sid = speciesid, prob = probability) %>%
  filter(am_sid %in% spp_am_v_iucn$am_sid)
### length(unique(am_spp_cells$am_sid)) results in 2314 species found
write_csv(am_spp_cells, file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'))

@@

```{r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### Create files for species list with area info, both a complete list and a trimmed list

@@

data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')
reload <- FALSE
if (!file.exists(data_file) | reload) {

@@

  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))
  create_spp_list2(spp_list_base, data_file, am_spp_cells, iucn_spp_cells, loiczid_raster)

@@

spp_list <- spp_list %>%
  mutate(sm_perc = ifelse(is.na(sm_perc), 0, sm_perc)) %>%
  filter(!area_iucn == 0)
  ### will ditch the ones with no IUCN area - formerly caused by DD species
  ### being filtered out.  Now, none suffer this consequence.

>>>>

```{r verify_spp_names_in_shps, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

# x <- spp_iucn_maps %>%
#   filter(is.na(iucn_subpop) & !is.na(iucn_sid)) %>%
#   filter(! spp_group %in% c('REPTILES', 'SEASNAKES', 'MANGROVES', 'SEAGRASSES')) %>%
#   filter(sciname %in% .$sciname[duplicated(.$sciname)])
### 37 instances of duped scinames with different IDs

api_key <- scan(file.path(dir_anx, 'api_key.csv'), what = 'character')

spp_id_url <- 'http://apiv3.iucnredlist.org/api/v3/species/id/%s?token=%s'

sid_list <- spp_iucn_maps %>%
  filter(!is.na(iucn_sid)) %>% 
  .$iucn_sid %>%
  unique() %>%
  sort()

ptm <- proc.time()
spp_names_df <- mc_get_from_api(spp_id_url, sid_list, api_key, delay = 1, cores = 16) %>%
  select(iucn_sid = taxonid, sciname2 = scientific_name)
proc.time() - ptm

spp_iucn_maps1 <- spp_iucn_maps %>%
  left_join(spp_names_df, by = 'iucn_sid') %>% ### here, both the shapefile name and API name
  select(-sciname) %>%
  rename(sciname = sciname2) ### now just the API name by ID number

write_csv(spp_iucn_maps1, file.path(dir_data, 'spp_iucn_maps_verified.csv'))

@@

spp_iucn_maps <- read_csv(file.path(dir_data, 'spp_iucn_maps_verified.csv'))

@@

spp_all <- spp_iucn_maps %>%
  filter(iucn_sid %in% spp_am_maps$iucn_id) %>%
  left_join(spp_am_maps %>% 
              rename(am_sciname = sciname) %>%
              mutate(iucn_sid = iucn_id), ### keep the IUCN noted ID separate from AM-noted IUCN id
            by = c('iucn_sid')) %>%
  bind_rows(x <- spp_iucn_maps %>%
              filter(!iucn_sid %in% spp_am_maps$iucn_id) %>%
              full_join(spp_am_maps %>%
                          mutate(am_sciname = sciname) %>%
                          filter(!iucn_id %in% spp_iucn_maps$iucn_sid | is.na(iucn_id)),
                        by = c('sciname'))) %>%
  distinct()

# x <- spp_all %>% 
#   filter(is.na(iucn_subpop) & !is.na(iucn_sid)) %>%
#   filter(iucn_sid %in% .$iucn_sid[duplicated(.$iucn_sid)])
### just seasnakes/reptiles and mangroves/seagrasses

# length(unique(spp_all$iucn_sid)) ###  4028, incl NA
# length(unique(spp_all$am_sid))   ### 22890, incl NA
# qwer <- spp_all %>% filter(!is.na(iucn_sid) & !is.na(am_sid))
# length(unique(qwer$iucn_sid)) ### 2343
# length(unique(qwer$am_sid)) ### 2337

# 
# x <- spp_iucn_maps %>%
#   rename(iucn_sciname = sciname) %>%
#   full_join(spp_am_maps %>%
#               rename(am_sciname = sciname),
#             by = c('iucn_sid' = 'iucn_id')) %>%
#   select(iucn_sid, iucn_sciname, am_sid, am_sciname) %>%
#   mutate(name_match = iucn_sciname == am_sciname) %>%
#   distinct()
# asdf <- x %>% filter(!is.na(iucn_sid) & !is.na(am_sid)) %>%
#   filter(!is.na(iucn_sciname) & !is.na(am_sciname))
# length(unique(asdf$iucn_sid)) # 1880
# length(unique(asdf$am_sid)) # 1880
### 1880 species matched if only using ID number...

@@

# length(unique(spp_am_v_iucn$sciname)); length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))
### NOTE: 2332 unique scientific names, unique am_sid and unique iucn_sid.
### Seasnakes are included in Reptiles (IUCN).  Two "mangroves" that are actually seagrasses.
spp_am_v_iucn <- spp_am_v_iucn %>%
  group_by(am_sid, iucn_sid) %>%
  mutate(dupes = n() - 1) %>%
  ungroup() %>%
  filter(!(dupes == 1 & spp_group == 'MANGROVES')) %>%
  filter(spp_group != 'SEASNAKES') %>%
  select(-dupes)
# length(unique(spp_am_v_iucn$sciname)); length(unique(spp_am_v_iucn$am_sid)); length(unique(spp_am_v_iucn$iucn_sid))
### still same scinames and am_sid

@@

# spp_am_v_iucn <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))

@@

# iucn_spp_cells_files <- list.files(file.path(dir_anx, 'vAM_IUCN/iucn_intersections'), full.names = TRUE)
# iucn_spp_cells_all <- lapply(iucn_spp_cells_files, FUN = function(x) {
#   read_csv(x) %>%
#     mutate(iucn_sid = as.integer(iucn_sid))
#   }) %>%
#   bind_rows() %>%
#   setNames(tolower(names(.)))
# 
# spp_iucn_maps1 <- read_csv(file.path(dir_data, 'spp_iucn_maps_verified.csv')) %>%
#   select(sciname, iucn_sid_ver = iucn_sid) %>%
#   distinct()

### because the 2015 shapefiles for reptiles and mammals were modified
### to attach subpop IDs rather than parent IDs to subpops, we will undo
### this here by overwriting the IDs based on parent IDs.
# iucn_spp_cells_verified <- iucn_spp_cells_all %>%
#   left_join(spp_iucn_maps1, by = 'sciname') %>%
#   mutate(iucn_sid = iucn_sid_ver)
# 
# iucn_spp_cells <- iucn_spp_cells_verified %>%
#   filter(iucn_sid %in% spp_am_v_iucn$iucn_sid) %>%
#   filter(presence != 5) %>%
#   select(-iucn_sid_ver)

## length(unique(iucn_spp_cells$iucn_sid)) #  2330 species found...
##   some species with no cells reported.
## spp_missing <- spp_am_v_iucn %>% filter(!iucn_sid %in% iucn_spp_cells$iucn_sid)
##   2 spp with no cells reported.  Shapefile errors?
# write_csv(iucn_spp_cells, file.path(dir_anx, 'vAM_IUCN/iucn_spp_cells_2015.csv'))
# 
# message('Loading AquaMaps species to cell table.  This might take a few minutes.')
# am_spp_cells_all <- fread(file.path(dir_data_remote, 'aquamaps/d2015/csv/hcaf_sp_native_trunc.csv'),
#                          showProgress = TRUE)
# am_spp_cells <- am_spp_cells_all %>%
#   rename(am_sid = speciesid, prob = probability) %>%
#   filter(am_sid %in% spp_am_v_iucn$am_sid)
# ### length(unique(am_spp_cells$am_sid)) # results in 2332 species found
# write_csv(am_spp_cells, file.path(dir_anx, 'vAM_IUCN/am_spp_cells_2015.csv'))

@@

``` {r get_spp_list, echo = FALSE, warning = FALSE, message = FALSE, eval = TRUE}
### Create files for species list with area info, both a complete list and a trimmed list

@@

data_file <- file.path(dir_data, 'spp_list_w_area_all.csv')
reload <- TRUE
if (!file.exists(data_file) | reload) {

@@

  ### species info list - scinames, IUCN species ID, AM species ID, category etc
  spp_list_base <- read_csv(file.path(dir_data, 'spp_am_v_iucn.csv'))
  create_spp_list(spp_list_base, data_file, am_spp_cells, iucn_spp_cells, loiczid_raster)

@@

spp_list <- spp_list %>%
  mutate(sm_perc = ifelse(is.na(sm_perc), 0, sm_perc)) %>%
  filter(!area_iucn == 0)
  ### will ditch the ones with no IUCN area - formerly caused by DD species
  ### being filtered out.  Now, only 2 suffer this consequence.
